---
title: "How to import and use Ingenuity IPA enrichment data"
output:
  rmarkdown::html_vignette:
    df_print: kable
vignette: >
  %\VignetteIndexEntry{How to import and use Ingenuity IPA enrichment data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse=TRUE,
  comment="#>"
)
```

```{r setup}
library(multienrichjam);
library(jamba);
library(colorjam);
suppressPackageStartupMessages(library(ComplexHeatmap));
options("stringsAsFactors"=FALSE, "warn"=-1);
knitr::opts_chunk$set(
   fig.height=10,
   fig.width=10,
   fig.align="center"
)
ragg_png = function(..., res = 192) {
  ragg::agg_png(..., res = res, units = "in")
}
knitr::opts_chunk$set(dev = "ragg_png", fig.ext = "png")
```

## Import and use Ingenuity IPA enrichment data

This document describes steps recommended for using Ingenuity
IPA enrichment data.

Ingenuity IPA enrichment data can be exported using a function
`"Export All"` which by default creates one text file,
concatenating each enrichment table into one large file.

This workflow demonstrates the import process using two
IPA enrichment files used by Reese et al 2019
https://doi.org/10.1016/j.jaci.2018.11.043
to compare enrichment results in newborns to older children.

It therefore requires IPA enrichment results have already
been exported in text format from IPA.

### Import IPA data from text files

To import an IPA text file, use `importIPAenrichment()`:

```{r, import_ipa_1}
newborn_txt <- system.file("extdata",
   "Newborns-IPA.txt",
   package="multienrichjam");
newborn_dfl <- importIPAenrichment(newborn_txt);
```

The result is a list of `data.frame` objects, where each
`data.frame` represents one enrichment test. A convenient
way to see the dimensions of each `data.frame` is with
the function `jamba::sdim()`:

```{r, ipa_1_sdim}
sdim(newborn_dfl);
```

For MultiEnrichMap, we typically want to analyze multiple
IPA enrichment files, so we can wrap the call in
an `lapply()` function:

```{r, import_ipa_1_2}
newborn_txt <- system.file("extdata",
   "Newborns-IPA.txt",
   package="multienrichjam");
olderchildren_txt <- system.file("extdata",
   "OlderChildren-IPA.txt",
   package="multienrichjam");
ipa_files <- c(Newborns=newborn_txt,
   OlderChildren=olderchildren_txt)

ipa_l <- lapply(ipa_files, importIPAenrichment);
```

Now we can check the dimensions within each list using
`jamba::ssdim()`:

```{r, ipa_ssdim}
ssdim(ipa_l);
```

In most cases, each IPA file should contain the same
enrichment tests, for example `"Canonical Pathways"`,
`"Upstream Regulators"`, `"Diseases and Bio Functions"`,
etc. However, it is not always the case, so it is
recommended to check and verify each IPA file contains
at least the enrichment tests needed for downstream analysis.

## Analyze IPA enrichments from one enrichment test

IPA performs multiple enrichment tests, which are
done independently and with unique assumptions
and caveats. Therefore, I recommend using one
enrichment test at a time in MultiEnrichMap.

Extract one `data.frame` from each result:

```{r, extract_canonical}
library(igraph)
## Take only the Ingenuity Canonical Pathways
enrichList_canonical <- lapply(ipa_l, function(i){
   i[["Canonical Pathways"]];
});
sdim(enrichList_canonical);

## Convert data.frame to enrichResult
## multienrichjam::enrichDF2enrichResult
er_canonical <- lapply(enrichList_canonical, function(i){
   enrichDF2enrichResult(i,
      keyColname="Name",
      pvalueColname="P-value",
      geneColname="geneNames",
      geneRatioColname="Ratio",
      pvalueCutoff=1)
});
sdim(er_canonical);
kable_coloring(
   head(as.data.frame(er_canonical[[1]])),
   caption="Top 10 rows of enrichment data",
   row.names=FALSE) %>%
   kableExtra::column_spec(column=seq_len(ncol(er_canonical[[1]])),
      border_left="1px solid #DDDDDD",
      extra_css="white-space: nowrap;")

```

### run multiEnrichMap()

Now given a list of `data.frame` results, we can
run `multiEnrichMap()`:

```{r, mem_canonical, fig.height=9, fig.width=10}
mem_canonical <- multiEnrichMap(er_canonical,
   enrichBaseline=1,
   cutoffRowMinP=0.05,
   colorV=c("purple", "orange"),
   topEnrichN=20)
```

The output `mem_canonical` is a list containing various
results.

```{r, mem_sdim}
kable_coloring(
   sdim(mem_canonical),
   caption="sdim(mem_canonical)") %>%
   kableExtra::column_spec(column=seq_len(4),
      border_left="1px solid #DDDDDD",
      extra_css="white-space: nowrap;")
```

### Create a "Mem Plot Folio" (new Apr2020)

The other analysis steps describe in this vignette are
valid options, however in recent
work with `multienrichjam` we have developed a new function
to create a small folio of relates plots that we call
a "mem plot folio", and the function is `mem_plot_folio()`.

The `mem_plot_folio()` function creates the following plots by default:

1. Enrichment P-value heatmap (optionally colored by group)
2. Gene-pathway incidence matrix, clustered by column and by row
3. Cnet plot including all collapsed pathway clusters

    * The first plot labels clusters by `LETTERS`, "A", "B", "C", "D", etc.
    * The second plot labels clusters by the top `n` pathway names
    * The third plot labels clusters by the top `n` pathway names, but hides gene labels.
    This plot is best when there are too many genes to label.

4. Cnet plot using 1,2,3 exemplars per pathway cluster (configurable)

    * The first plot uses 1 exemplar per cluster.
    * The second plot uses 2 exemplars per cluster.
    * The third plot uses 3 exemplars per cluster.

5. Cnet plot of each pathway cluster individually

    * The number of plots is the number of heatmap pathway_clusters.

```{r, mem_folio, fig.height=10, fig.width=10}
mem_canonical_plots <- multienrichjam::mem_plot_folio(mem_canonical,
   pathway_column_split=4,
   node_factor=1,
   use_shadowText=TRUE,
   label_factor=1.2,
   do_which=c(1:5),
   verbose=TRUE,
   main="Canonical Pathways");
```

The object returned `mem_canonical_plots` will contain a list
of the graphical objects for each plot requested. For example,
each Cnet plot returns an `igraph` object which can be customized.


#### Commentary on the "Plot Folio"

In practice, we rarely found much benefit from the multi-enrichment map,
which shows pathways connected to pathways based upon Jaccard overlap.
This network view was often disorganized, not clearly clustered,
and lacked ability to see which genes drive the overlaps between
pathways. Perhaps it works best for the highly structured Gene Ontology (GO),
but in our hands GO was just not insightful for the broad range of
experiments we were analyzing.

Instead, we almost always gravitated toward the Cnet plot, a clever idea
by Dr. Guangchang Yu that connects pathways to genes, where genes also connect
naturally to other pathways. This plot for many of our collaborators has
seemed more intuitive, and answers the next question people often have:
"What are the shared genes?" Our subtle addition to this plot is to
color-code genes by the enrichment experiment, so you can see which genes
are shared or unique by experiment.

Since most enrichment results have far more pathways than can be shown
in one Cnet plot, we tried to prioritize which pathways to display.
The logic in Dr. Yu's `enrichplot` is to take the top `n` pathways by P-value,
however we noticed for our data this logic often shows redundant pathways and
misses orthologous biological functions, which might be enriched
to a lesser degree. So we started clustering pathways by P-value,
then by using the pathway-gene incidence matrix.

The pathway-gene incidence matrix worked best in our hands for this
clustering, partly because it clusters pathways by the genes involved,
and not by the pattern of enrichment across the enrichment tests
(which may not be related to the underlying genes.) So we cluster
pathways, then choose an arbitrary number of clusters, and took
one "exemplar" from each cluster to represent each cluster. This
practice is mostly effective, especially when the clusters are well-defined,
and when members of each cluster have relatively similar function.
We can increase the exemplars per cluster as needed. This strategy
is helpful because each pathway is very clearly labeled.

Most recently, we experimented with collapsing each pathway_cluster,
so each pathway_cluster includes all genes implicated by pathways
contained in the pathway_cluster. The benefit here is that no genes are lost
by selecting a subset of "exemplar" pathways -- all genes are retained
per cluster. Then instead of plotting pathway-gene network, we plot
a pathway_cluster-gene network. Each pathway_cluster is labeled
using the top `n` pathway names -- which is configurable, and is
only a visible label. Each pathway_cluster contains all the
pathways in the cluster.


### Directly create each individual plot

The plots in `mem_plot_folio()` are also available as individual
functions.

In general, it is most convenient to call `mem_plot_folio()` and
use the output `list` to create custom plots, however the functions
are also available to be called independently.

#### Plot the MultiEnrichMap network

We can view the "Multi Enrichment Map" itself
with `mem_multienrichplot()`.

This network connects pathways when they meet a Jaccard
overlap coefficient threshold based upon the shared genes
between the pathways. The default overlap is stored
by `multiEnrichMap()` in the output object `mem_canonical`.


```{r, mem_plot_1}
g <- mem_multienrichplot(mem_canonical,
   do_plot=FALSE,
   overlap=0.3,
   node_factor=2,
   repulse=3.5)
jam_igraph(g,
   node_factor=2,
   use_shadowText=TRUE)

```

The overlap threshold can be customized manually,
or through the helper function `mem_find_overlap()`.
This function attempts to guess a reasonable overlap
threshold based upon the network connectivity,
by optimizing the cluster size and total nodes involved.

```{r, mem_plot_1b}
g <- mem_multienrichplot(mem_canonical,
   overlap=mem_find_overlap(mem_canonical),
   do_plot=FALSE,
   node_factor=3,
   repulse=3.5)
jam_igraph(g,
   node_factor=3,
   use_shadowText=TRUE)
```


Notice there are distinct subnetworks, called "components",
which are not connected to each other. A useful follow-up
technique is to graph one component at a time, which we do
using the helper function `subset_igraph_components()`.
Components are ordered by size, largest to smallest,
so you can keep the largest using argument `keep=1`,
or the second largest with `keep=2`, and so on.

We also call two other helper functions:

1. `removeIgraphBlanks()` removes blank colors from multi-color
nodes, such as pie nodes, or colored rectangle nodes. It helps
show only the remaining colors without the whitespace.
2. `relayout_with_qfr()` applies Fruchterman-Reingold layout
to an `igraph` object. It's convenient to adjust `repulse` here,
which is the recommended method for adjusting the spacing
between nodes. This function also updates other useful
attributes such as the node label angle, which makes labels
appear offset away from the majority of edges. (It helps
more with Cnet plots below.)

```{r, mem_plot_2}
## You can alternatively pull out any other component
g_sub <- subset_igraph_components(g, keep=1);

## Re-apply network layout, and remove blank colors
g_sub <- relayout_with_qfr(repulse=3.5,
   removeIgraphBlanks(g_sub))

## Plot
jam_igraph(g_sub,
   node_factor=3,
   use_shadowText=TRUE)
```



#### Plot the enrichment P-value matrix

A useful overview of the enrichment results, including
the overlaps across datasets, is shown in a heatmap
of `-log10(Pvalue)`. The helper function
`mem_enrichment_heatmap()` is used here.

The argument `p_cutoff` is used to set the Pvalue
below which cells are colorized -- every P-value above
this threshold is not colored, and displayed as white,
even when the P-value is less than 1.

```{r, enrich_hm}
mem_enrichment_heatmap(mem_canonical,
   p_cutoff=0.05);
```

The same data can be plotted as a heatmap.

```{r, enrich_hm_1}
mem_enrichment_heatmap(mem_canonical,
   style="heatmap",
   p_cutoff=0.05);
```


The argument `color_by_column=TRUE` will optionally apply
a color gradient to each column, using the same P-value
threshold defined by `p_cutoff`. In this case, cells are
labeled to indicate the actual enrichment P-value.

```{r, enrich_hm_2}
mem_enrichment_heatmap(mem_canonical,
   style="heatmap",
   color_by_column=TRUE);
```


#### Plot the pathway-gene incidence matrix

We can view the pathway-gene matrix using the function
`mem_gene_path_heatmap()`. This function uses data in the
`mem` object: `"memIM"` which contains the gene-pathway
incidence matrix, `"geneIM"` which contains the gene-dataset
incidence matrix, and `"enrichIM"` which contains the
pathway-dataset matrix.

This function attempts to "guess" a reasonable number
of column and row splits -- which breaks the column and
row dendrograms into separate sub-clusters. These
breaks are visual only, but can be useful in follow-up
analyses shown later. You can override these numbers
with arguments `column_split` and `row_split`.

Also, the colors used across the top of the heatmap indicate
the enrichment P-value, and are intentionally scaled so all
P-values above the P-value threshold are not colored, and
are therefore white.

This function uses the amazing `ComplexHeatmap::Heatmap()`
function, which enables many awesome customizable features.
The `...` argument is passed to `ComplexHeatmap::Heatmap()`,
so you can use those features as needed.

```{r, mem_hm}
hm <- mem_gene_path_heatmap(mem_canonical,
   column_cex=0.5,
   row_cex=0.6);
hm;
```

As a follow-up analysis, you can pull out each pathway
cluster from the heatmap itself, using `heatmap_column_order()`:

```{r, mem_hm_sets}
hm_sets <- heatmap_column_order(hm);
hm_sets;
```

### Plot the Pathway-Gene Concept network (Cnet plot)

We can view the Cnet plot (concept network plot) which
shows the pathway-gene relationship.

The helper function `memIM2cnet()` creates a Cnet plot
from the `mem_canonical` output. Here, we also pipe the
result through other helper functions:

* `fixSetLabels()` applies pathway label word wrap
* `relayout_with_qfr()` applies network layout and adjusts node labels
* `removeIgraphBlanks()` removes blank colors from the `igraph` nodes

```{r, cnet_plot_1}
#cnet <- mem_canonical$multiCnetPlot1b;
cnet <- mem_canonical %>% 
   memIM2cnet() %>%
   fixSetLabels() %>%
   removeIgraphBlanks() %>%
   relayout_with_qfr(repulse=4);
#plot(cnet, vertex.label.cex=0.6);
par("mar"=c(5,4,4,2)+0.1)
jam_igraph(cnet,
   use_shadowText=TRUE,
   node_factor=0.5,
   vertex.label.cex=0.6);
mem_legend(mem_canonical);
```

You can extract the largest connected subnetwork to plot, as before.

```{r, cnet_plot_2}
g2 <- cnet;
g2_sub <- subset_igraph_components(cnet, keep=1)

#plot(g2_sub);
jam_igraph(g2_sub,
   use_shadowText=TRUE,
   label_factor=0.5,
   node_factor=0.5);
```

#### Customizing the Cnet plot

The most useful customization is to subset the pathway
nodes.

One useful method to choose a subset of pathways 
is to view one heatmap cluster at a time. Here
we will use `hm_sets` made previously.

```{r, cnet_plot_1_custom}
cnet_sub <- subsetCnetIgraph(cnet,
   repulse=3.5,
   includeSets=unlist(hm_sets[c("A")]));
jam_igraph(cnet_sub,
   node_factor=1,
   use_shadowText=TRUE,
   label_dist_factor=3,
   label_factor=1.3);
mem_legend(mem_canonical);
```


#### Subset the Pathway-Gene Cnet plot

You can also subset the Pathway-Gene Cnet plot using
specific names, or by some network criteria, using
the function `subsetCnetIgraph()`.

Pathways are subsetted to require at least 6 genes,
using the argument `minSetDegree`,
which can be useful to help simplify the graph structure.

Similarly, the `minGeneDegree=2` argument can be used to
display only genes present in 2 or more pathways, but for
now we will not use that argument.

During the subset operation, some convenient default operations
are also performed, controlled by function arguments:

* singlet nodes (with no connection to any other) are removed,
`remove_singlets=TRUE`
* the layout is re-calculated, `force_relayout=TRUE`
* nodes are re-ordered by color, `do_reorder=TRUE`
* labels are re-positioned to have proper angle offset away
from the edges entering each node, `spread_labels=TRUE`
* blank colors are not removed from pie and colored rectangle
nodes, `remove_blanks=FALSE`, but can be during this step by
`remove_blanks=TRUE`.

```{r, subnet, fig.height=8, fig.width=7}
cnet3 <- multienrichjam::subsetCnetIgraph(cnet,
   repulse=5,
   minSetDegree=6,
   minGeneDegree=1);
jam_igraph(cnet3,
   node_factor=0.7,
   use_shadowText=TRUE);
mem_legend(mem_canonical);
```

#### Custom igraph plot function jam_igraph()

The `jam_igraph()` function is a custom version of `igraph::plot()`
intended to enhance the base function in a few ways:

* `edge_bundling="connections"` improves the rendering of edges
by bundling edges from node clusters, so they are drawn with
a bezier curve
* `use_shadowText=TRUE` will draw labels with a contrasting border
to improve legibility of text labels
* `rescale=FALSE` keeps the network layout aspect ratio
instead of scaling the coordinates to fit the size
of the plot window. It also properly scales the node and edge sizes.
* new arguments to help resize existing values:

   * `label_factor` re-scales the `label.cex` values by a multiplier
   * `node_factor`, `edge_factor` re-scales the `node.size` and `edge.width`
   by a multiplier
   * `label_dist_factor` re-scales the `label.dist` values by a multiplier
   
The function below replicates the previous Cnet plot, with
these changes:

* `node_factor=2` makes the nodes 2x larger
* `label_factor=1.2` makes the `label.cex` values 1.2x larger
* `label_dist_factor=3` makes the label distance from the node center
3x larger, which emphasizes the label placement at an angle away from
the edges of each node.

```{r, jam_igraph_1, fig.height=8, fig.width=7}
jam_igraph(cnet3,
   edge_factor=2,
   node_factor=1.2,
   label_factor=1.2, 
   use_shadowText=TRUE,
   label_dist_factor=3);
```

One more fancy alternative, you can colorize the edges based
upon the node colors, a great idea inspired by the Gephi
netowrk visualization tool.

```{r, jam_igraph_2, fig.height=8, fig.width=7}
jam_igraph(color_edges_by_nodes(cnet3, alpha=0.7),
   edge_bundling="connections",
   edge_factor=2,
   node_factor=1,
   label_factor=1.2, 
   use_shadowText=TRUE,
   label_dist_factor=3)
```

